---
title: "GBIF rgbif demo"
author: "Dag Endresen, GBIF.no, http://orcid.org/0000-0002-2352-5497"
date: "February 2, 2018"
output: html_document
---

# Clear workspace (be careful)
#rm(list = ls())

# Set working directory, here: to the same directory as the RMD-script
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
#getwd() ## will display the working directory
#dir.create(file.path("./demo_files")) ## create a folder for demo files

# Some useful R-packages to install, that we will use
#install.packages("rmarkdown")
#install.packages("rgbif")
#install.packages("maps")
#install.packages("mapproj")
#install.packages("raster")

# Load R-packages
#Use library(_pk-name_) or require(_pk-name_) to load packages
require(rgbif) # r-package for accessing GBIF
require(maps)
require(mapproj)
require(mapdata)
require(maptools) # mapping tools for spatial objects
require(rgdal) # provides the Geospatial Data Abstraction Library
require(raster) # spatial raster data management, works well with dismo

# Example taxon liverleaf (blaaveis:no)
```{r message=FALSE}
sp_name <- "Hepatica nobilis" ## Choose your own species
#sp_name <- "Hordeum vulgare" ## Choose your own species
key <- name_backbone(name=sp_name, kingdom="Plantae")$speciesKey
```

# `r sp_name`

# Example GBIF data from Norway
```{r echo=FALSE, message=FALSE}
sp <- occ_search(taxonKey=key, return="data", hasCoordinate=TRUE, country="NO", limit=500) 
gbifmap(sp, region = "norway")
```

# Example global species occurrence data from GBIF
```{r message=FALSE}
require(rgbif) # r-package for accessing GBIF
sp <- occ_search(taxonKey=key, return="data", hasCoordinate=TRUE, limit=1000) 
gbifmap(sp)
```

## Extract coordinates for smaller array
```{r}
xy <- sp[c("decimalLongitude","decimalLatitude")] ## Extract only the coordinates
sp_xy <- sp[c("species", "decimalLongitude","decimalLatitude")] ## Input format for Maxent
```

## Write dataframe to file (useful for Maxent etc)
```{r}
#write.table(sp_xy, file="./sp_xy.txt", sep="\t", row.names=FALSE, qmethod="double") ## Input for Maxent
```

## Read data file back into R
```{r}
#rm(sp_xy) ## remove/delete vector sp_xy from the environment, before re-loading
#sp_xy <- read.delim("./sp_xy.txt", header=TRUE, dec=".", stringsAsFactors=FALSE)
```


## Read environment layer into R
```{r include=FALSE}
require(raster) # spatial raster data management, works well with dismo
env <- getData('worldclim', var='bio', res=10) # 10 degree resolution (approx 18 km)
#env <- getData('worldclim', var='bio', res=2.5) # 2 degree 5 minute resolution (approx 5 km)
```

## Plot environment layers and species occurrences on a map
```{r fig.cap="Figure: GBIF data plotted on environment layer map", echo = FALSE}
#plot(env, 1) # plot the first bioclim layer (BIO1 = Annual Mean Temperature)
plot(env, 12, main=NULL, axes=FALSE) # plot bioclim layer, BIO12 = Annual Precipitation
#title(main = "BioClim 12, Annual precipitation")
title(main = bquote(italic(.(sp_name)) ~occurrences~plotted~on~annual~precipitation))
axis(side=2, tck = -0.04, col.ticks="gray") ## add axis only left
points(xy, col='blue') # plot species occurrence points to the map
```

```{r echo = FALSE}
plot(env, 1, main=NULL, axes=FALSE)
title(main = bquote(italic(.(sp_name)) ~occurrences~":"~on~annual~mean~temperature))
points(xy, col='red', pch=20) # plot species occurrence points to the map (smaller dots)
```


```{r}
```


### This script follows R Markdown, http://rmarkdown.rstudio.com/
