---
title: "Workshop intro and some R setup examples/notes"
author: "Dag Endresen (GBIF.no) https://orcid.org/0000-0002-2352-5497"
date: "February 3, 2018"
output: github_document
---


# Nordic Oikos 2018 - R workshop

Scientific reuse of openly published biodiversity information: Programmatic access to and analysis of primary biodiversity information using R. Nordic Oikos 2018, pre-conference R workshop, 18 and 19 February 2018. Further information [here](http://www.gbif.no/events/2018/Nordic-Oikos-2018-R-workshop.html).


### Set working directory, here: to the same directory as the RMD-script
```{r include=FALSE}
# setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) ## does not work inside Rmd (?)
# however, at least knitr seems to assume working directory same as the script anyway
getwd() ## will display the working directory
```

### Create a folder for demo data
```{r include=FALSE}
## Note however, that RMarkdown with knitr works best with all files in the same directory
#dir.create(file.path("./demo_data")) ## create a folder for demo files (inside working directory)
#dir.create(file.path("../demo_data")) ## create a folder for demo files (next to working directory)
#dir.create(file.path("..", "demo_data")) # will work better accross OS? on Windows?
```

### Some useful R-packages to install, that we will use during the workshop
```{r}
## It is possible to install R-packages from the script
#install.packages("rmarkdown")
#install.packages("rgbif")
#install.packages("maps")
#install.packages("mapproj")
#install.packages(mapdata)
#install.packages("raster")
```

### Load R-packages
```{r}
# Use functions library() or require() to load the R-packages you need

#require(rgbif) # r-package for accessing GBIF
#require(maps)
#require(mapproj)
#require(mapdata)
#require(maptools) # mapping tools for spatial objects
#require(rgdal) # provides the Geospatial Data Abstraction Library
#require(raster) # spatial raster data management
#library(gapminder)
#library(tidyverse) ## includes dplyr etc for handling data, select(), filter()
```

### Clear workspace (be careful)
```{r}
# Large arrays and dataframes can consume huge parts of your working memory,
# and you probably want to remove some of them from memory when they are not needed anymore.

#rm(list = ls()) # Be careful, this line cleans the entire R workspace
#rm(sp) # Probably better to remove large arrays and dataframes individually
```


## This script uses [R Markdown](http://rmarkdown.rstudio.com/)

[R Markdown](http://rmarkdown.rstudio.com/) is a convinient format for writing R-scripts, introduced in 2014. R Markdown integrates very well in RStudio (and can also be used form the command line). Session 2 will provide an introduction to R-scripting including R Markdown.

